---
title: "Class 2"
author: "SCOVIA Muhoza"
date: "2025-10-16"
output: html_document
---

## Class2_Cont


# Control structures
#Grouping

Control structures in R allow you to control the flow of execution of a series of R
 expressions. Basically, control structures allow you to put some “logic” into your R
 code, rather than just always executing the same R code every time. Control
 structures allow you to respond to inputs or to features of the data and execute
 different R expressions accordingly. 
 Commonly used control structures are •
 if and else: testing a condition and acting on it • for: execute a loop a fixed number of times
 • while: execute a loop while a condition is true • repeat: execute an infinite loop
 (must break out of it to stop) • break: break the execution of a loop • next: skip an
 interaction of a loop
 
 if (condition){ ## do something } ## continue with the rest of the code.
 if( condition){ do something
 } else { do something else
 }
 You cana lso create a series of test by following the initial if with a number of else ifs
 if(condition){ do something
 }else if (condition 2){ Do something different }else{ do something different 
 
```{r}
 #Function on R
 f<-function(x) x^2
 formals(f)
 body(f)
 environment(f)
 
```
# ifelse(test,yes,No)

```{r}
 pvalues<-c(.867,0.0054,0.0018,0.1572,0.0183,0.5386)
 results<-ifelse(pvalues<0.05,"Significant","Notsignificant")
 results 

```

```{r}
x<-runif(1,0,10)
 if(x>3){
 y<-10
 }else{
 y<-0
 }
 #Thevalueofyisasetdependingonwhetherx>3ornot.Thiscanalsobeachieved by
 y<-if(x>3){
 10
 }else{
 0
 }
  for(i in 1:10){
 print(i)
 }
```
```{r}
 x<-c("a","b","c","d")
 x[3]

 for(i in 1:5){
 print(x[i])
 }

```
```{r}
for(i in 1:5)print(1:i)

```

```{r}
for(i in 5:1)print(1:i)
```
## whileLoops
 It begins by testing a condition,if it is true,then they execute the loop body.once the
 loop body is executed,the condition is tested again,until the condition is false.after
 which the loop exits.
 

```{r}
 count<-1
 while(count<10){
 print(count)
 count<-count+1
 }
```
```{r}
 z<-5
 set.seed(1)
 while(z>=3&&z<=10){
 coin<-rbinom(1,1,0.5)
 if(coin==1){
 z=z+1
 }else{
z<-z-1
 }
 }
 print(z)
```
## next,break

```{r}
#This is used to skip an iteration of aloop
 for(i in 1:100){
 print(1:i)
 if(i>20){
 break
 }
 }
```

# Function

```{r}

 functionname<-function(parameters){statementsreturn(value)}
 f1<-function(x)
 {result<-x^2+2
   return(result)}
 f2<-function(x,y)
 {result<-x2+y2-4
   return(result)}
 f<-function(x,y){
 result<-x+(2*y)+3
 return(result)
 }
f(2,3)

 f(2,3)
  #You can use args() function to view theparameter names and default values
```
Exercises
 Make functions that calculate summary statistics
 Make a function to calculate two sample t test
 Applying functions to matrices and dataframe.
 
```{r}
 a<-4
 sqrt(4)
```
```{r}
 b<-c(1,243,5.754,2.987)
 round(b)
 

```
```{r}
 c<-matrix(runif(12),nrow=3)
 c

```
```{r}
log(c)
```
```{r}
mean(c)
```
Notice that the mean of matrixc results in a scalar(0.444).the mean()take the average of all 12 elements in the matrix.But what if you want the three row means or the four column means?

 R provides a function ,apply() that allows to apply an arbitrary function to any dimension of a matrix,array or dataframe.The format for the apply() function is
apply(x, MARGIN, FUN,…) where x stands for the data object, Margin can be 1(rows)
 and 2(columns ) ##Col/Row Sums and Means
 
 • rowSums=apply(x, 1, sum) • rowMeans= apply(x, 1, mean)• colSums =apply(x, 2,
 sum) •colMeans= apply(x, 2, mean)


 
 
#lappy function
 lapply() takes three inputs:,x,a list,a function, and…, It applies to each element of the
 list and returns a new list. lapply(x,f,…). It is called function because it takes function
 as an argument. Assume we have a dataframe df. instead of assigningthe result pf
 lappliy() to df, we will assign them to df[] to ensure we get a data frame

```{r}
fix_missing<-function(x){
 x[x==-99]<-NA
 x
 }
 function1<-function(x){
 x^2
 }
 curve(function1,-4,4)

```
#df[]<-lappy(df,fix_missing)
 It works for any number of columns.There is no wayto accidentally miss a column
There is no way to accidentally treat one column  differently than  another
 It easy to generalize this technique to a subset of columns
 
#df[1:5]<-lapply(df[1:5],fix_missing)


# sapply

 sapply() and vapply(),variants of lapply()that produces vectors,matrices,and arrays as Output,instead of lists map(),and mapply which iterate over multiple input data structures in parallel Another important function when dealing with big data is split()-tapply

```{r}
#Titanic data
 str(Titanic)
```
```{r}
 View(Titanic)
 apply(Titanic,c(1,2),sum)
```

```{r}
 options(digits=2)
 apply(Titanic,c(1,2),sum)[3:4,]
```
```{r}
apply(Titanic,c(1,4),sum)[3:4,]
```

```{r}
 apply(Titanic,c(1,2,4),sum)
```

```{r}
#apply(Titanic,c(1,2,4),sum)[3:4,]
 apply(Titanic,c(1,2,4),sum)[3:4,,]
```

```{r}
ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])
```
```{r}
ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,]
```
```{r}
ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,]
```

```{r}
 prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,],margin=2)
```
```{r}
prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[3:4,],margin=2)
```

```{r}
matrix(prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,],margin=2),nrow=2,dimnames=list(dimnames(Titanic)$Survived,dimnames
 (Titanic)$Sex))
```

```{r}
 matrix(prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[3:4,],margin=2),nrow=2,dimnames=list(dimnames(Titanic)$Survived,dimnames
 (Titanic)$Sex))
```
# Createdataset
```{r}

 Student<-c("John ncuti","Angela bakame","Bruce wizeye","Alexis aganze","claude Rukundo","Joel Kagabo","Mary ineza")
 Math<-c(600,412,358,495,512,410,522)
 Science<-c(95,99,80,82,75,89,77)
 English<-c(25,22,18,20,29,30,27)
 roster<-data.frame(Student,Math,Science,English,stringsAsFactors=FALSE)
 
  ###standardize variables and obtains the performance scores because they are reported on  different scale(With widely differing means and standard deviations,we need to make them comparable before we combine them.)
```

```{r}
 z<-scale(roster[,2:4])
 ###performance of each students using row means and adding them to roster using cbind()
 score<-apply(z,1,mean)
 roster<-cbind(roster,score)
###Grades the students:quantile function gives thepercentile rank of each student's performance score check the cutoff of A
```

```{r}
 y<-quantile(score,c(.8,.6,.4,.2))
 ##create a grade variable us
 roster$grade[score>=y[1]]<-"A"
 roster$grade[score<y[1]&score>=y[2]]<-"B"
 roster$grade[score<y[2]&score>=y[3]]<-"C"
 roster$grade[score<y[3]&score>=y[4]]<-"D"
 roster$grade[score<y[4]]<-"F"
 ####Dealingwithnames
 name<-strsplit((roster$Student),"")
 lastname<-sapply(name,"[",2)
 firstname<-sapply(name,"[",1)
 roster<-cbind(firstname,lastname,roster[,-1])
 roster<-roster[order(lastname,firstname),]
 roster
```


#Aggregationandreshaping
#transpose
```{r}
 cars<-mtcars[1:5,1:4]
 cars
```
```{r}
t(cars)
```

# aggregate data
aggregate() collapse data in R using one or more by variables and a defined function

```{r}
 options(digits=3)
 attach(mtcars)
 aggdata<-aggregate(mtcars,by=list(cyl,gear),FUN=mean,na.rm=TRUE)
```

# Reshape
 Step 1: install the package reshape2 step 2: melt data step 3: Cast the melted data
 into any shape you desire
 During the cast, you can aggregate the data with any function you wish

```{r}
 id<-c(1,1,2,2)
 time<-c(1,2,1,2)
 x1<-c(5,3,6,2)
 x2<-c(6,5,1,4)
 mydata<-data.frame(id,time,x1,x2)
```

